Descriptive Complexity for Counting Complexity Classes

### 2017-03-07 LMCS/framework.tex
journam 7 mar

To define the permanent of a 0-1 matrix in $\qso$, assume that for a binary relation symbol $S$, $\op{permut}(S)$ is an $\fo$-formula that is true if and only if $S$ is a permutation, that is, a total bijective function (the definition of $\op{permut}(S)$ is straightforward).
%\martin{abreviar $\op{permutation}$ a $\op{permut}$ ahorra un par de lineas}
Then the following is a $\qso$-formula defining the permanent of a matrix:
\[
\alpha_3 := \sa{S} \op{permut}(S) \cdot \pa{x} (\ex{y} S(x,y) \wedge M(x,y)).
\]
Intuitively, the subformula $\beta(S) := \pa{x} (\ex{y} S(x,y) \wedge M(x,y))$ calculates the value $\prod_{i=1}^n A[i, \sigma(i)]$ whenever $S$ encodes a permutation $\sigma$.
Moreover, the subformula $\op{permut}(S) \cdot \beta(S)$ returns $\beta(S)$ when $S$ is a permutation, and returns $0$ otherwise (i.e. $\op{permut}(S)$ behaves like a filter). 
Finally, the second order sum aggregates these values iterating over all binary relations and calculating the permanent of the matrix.
We would like to finish with this example by highlighting the similarity of $\alpha_3$ with the permanent formula. 
Indeed, an advantage of $\qso$-formulae is that the first- and second-order quantifiers in the quantitative level naturally reflect the operations used to define mathematical formulae.
\end{exa}

>>>>

To define the permanent of a 0-1 matrix in $\qso$, assume that for a binary relation symbol $S$, $\op{permut}(S)$ is an $\fo$-formula that is true if and only if $S$ is a permutation, that is, a total bijective function (the definition of $\op{permut}(S)$ is straightforward).
%\martin{abreviar $\op{permutation}$ a $\op{permut}$ ahorra un par de lineas}
Then the following is a $\qso$-formula defining the permanent of a matrix:
\[
\alpha_3 := \sa{S} \op{permut}(S) \cdot \pa{x} (\ex{y} S(x,y) \wedge M(x,y)).
\]
Intuitively, the subformula $\beta(S) := \pa{x} (\ex{y} S(x,y) \wedge M(x,y))$ calculates the value \linebreak $\prod_{i=1}^n A[i, \sigma(i)]$ whenever $S$ encodes a permutation $\sigma$.
Moreover, the subformula $\op{permut}(S) \cdot \beta(S)$ returns $\beta(S)$ when $S$ is a permutation, and returns $0$ otherwise (i.e. $\op{permut}(S)$ behaves like a filter). 
Finally, the second order sum aggregates these values iterating over all binary relations and calculating the permanent of the matrix.
We would like to finish with this example by highlighting the similarity of $\alpha_3$ with the permanent formula. 
Indeed, an advantage of $\qso$-formulae is that the first- and second-order quantifiers in the quantitative level naturally reflect the operations used to define mathematical formulae.
\end{exa}

### 2017-03-07 LMCS/proof-qe0-fp-qe1-totp-fptras.tex
journam 7 mar

\begin{lemma} \label{conj-mult}
	If $\LL$ is a fragment closed under conjunction, then $\eqso(\LL)$ is closed under binary multiplication.
\end{lemma}
\begin{proof}
	We will define a recursive function $\tau$ that receives a formula $\alpha$ over the grammar of $\eqso(\LL)$ extended by binary product, and outputs an equivalent formula $\tau(\alpha)$ over the unextended grammar of $\eqso(\LL)$. In fact, the formula $\tau(\alpha)$ is in $\LL$-SNF. First we replace each constant $s$ in $\alpha$ for $(\top \add \cdots \add \top)$ ($s$ times). If $\alpha = \varphi$, then we define $\tau(\alpha) = \alpha$. We assume that for every $\beta$ that has less algebraic operators than $\alpha$, $\tau(\beta)$ is in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$ then we define $\tau(\alpha) = \tau(\alpha_1) + \tau(\alpha_2)$. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$, then we define $\tau(\alpha)$ as the formula in $\LL$-SNF that is equivalent to $\sa{x}\tau(\beta)$ and to $\tau(\alpha) = \sa{X}\tau(\beta)$, respectively. If $\alpha = (\alpha_1 \cdot \alpha_2)$, we assume that each $\alpha_i$ is in $\LL$-SNF. We identify three cases. (1) Some $\alpha_i$ is equal to $\sum_{j = 1}^n\beta_j$ for $n > 1$. Suppose wlog. that it is $\alpha_1$. We then define $\tau(\alpha) = \sum_{j = 1}^n\tau(\beta_j\cdot\alpha_2)$. In the following cases, $\alpha_1$ and $\alpha_2$ are in $\LL$-SNF. (2) If some $\alpha_i$ is equal to $\sa{X}\beta$ or $\sa{x}\beta$, we define $\tau(\alpha)$ as the $\LL$-SNF formula that is equivalent to $\sa{x}\tau(\beta\cdot\alpha_2)$ and $\sa{X}\tau(\beta\cdot\alpha_2)$, respectively. The remaining case is (3) $\alpha_1 = \varphi_1$ and $\alpha_2 = \varphi_2$ where each $\varphi$ is an $\LL$ formula. Then we define $\tau(\alpha) = \varphi_1 \wedge \varphi_2$. This covers all possible cases for $\alpha$. For every pair of formulas $\alpha,\beta$ in $\eqso(\LL)$, we have that their multiplication $(\alpha\cdot\beta)$ is a formula in the grammar $\eqso(\LL)$ extended by binary product, and so, there exists an equivalent formula $\tau(\alpha\cdot\beta)$ which is in unextended $\eqso(\LL)$.
\end{proof}
Since $\loge{1}$ is closed under conjunction, this also holds for $\eqso(\loge{1})$. This concludes the proof.

>>>>

\begin{lem} \label{conj-mult}
	If $\LL$ is a fragment closed under conjunction, then $\eqso(\LL)$ is closed under binary multiplication.
\end{lem}
\proof
	We will define a recursive function $\tau$ that receives a formula $\alpha$ over the grammar of $\eqso(\LL)$ extended by binary product, and outputs an equivalent formula $\tau(\alpha)$ over the unextended grammar of $\eqso(\LL)$. In fact, the formula $\tau(\alpha)$ is in $\LL$-SNF. First we replace each constant $s$ in $\alpha$ for $(\top \add \cdots \add \top)$ ($s$ times). If $\alpha = \varphi$, then we define $\tau(\alpha) = \alpha$. We assume that for every $\beta$ that has less algebraic operators than $\alpha$, $\tau(\beta)$ is in $\LL$-SNF. If $\alpha = (\alpha_1 + \alpha_2)$ then we define $\tau(\alpha) = \tau(\alpha_1) + \tau(\alpha_2)$. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$, then we define $\tau(\alpha)$ as the formula in $\LL$-SNF that is equivalent to $\sa{x}\tau(\beta)$ and to $\tau(\alpha) = \sa{X}\tau(\beta)$, respectively. If $\alpha = (\alpha_1 \cdot \alpha_2)$, we assume that each $\alpha_i$ is in $\LL$-SNF. We identify three cases. (1) Some $\alpha_i$ is equal to $\sum_{j = 1}^n\beta_j$ for $n > 1$. Suppose wlog. that it is $\alpha_1$. We then define $\tau(\alpha) = \sum_{j = 1}^n\tau(\beta_j\cdot\alpha_2)$. In the following cases, $\alpha_1$ and $\alpha_2$ are in $\LL$-SNF. (2) If some $\alpha_i$ is equal to $\sa{X}\beta$ or $\sa{x}\beta$, we define $\tau(\alpha)$ as the $\LL$-SNF formula that is equivalent to $\sa{x}\tau(\beta\cdot\alpha_2)$ and $\sa{X}\tau(\beta\cdot\alpha_2)$, respectively. The remaining case is (3) $\alpha_1 = \varphi_1$ and $\alpha_2 = \varphi_2$ where each $\varphi$ is an $\LL$ formula. Then we define $\tau(\alpha) = \varphi_1 \wedge \varphi_2$. This covers all possible cases for $\alpha$. For every pair of formulas $\alpha,\beta$ in $\eqso(\LL)$, we have that their multiplication $(\alpha\cdot\beta)$ is a formula in the grammar $\eqso(\LL)$ extended by binary product, and so, there exists an equivalent formula $\tau(\alpha\cdot\beta)$ which is in unextended $\eqso(\LL)$.
\qed
Since $\loge{1}$ is closed under conjunction, this also holds for $\eqso(\loge{1})$. This concludes the proof.
